---
title: "Wisconsin breast cancer: logistic regression vs. XGBoost"
format: pdf
---

Three classifiers for `bdiag.csv`, predicting `diagnosis` (B/M) from the cell-nucleus measurements: a two-variable logistic regression, a full logistic regression, and an XGBoost model (state of the art on tabular data).

_The code to run and evaluate the three models was produced by Claude (Opus 4.6) and reviewed manually._

```{r}
library(xgboost)
library(pROC)

set.seed(49773858)
```

## Data

The CSV has a trailing comma that produces an empty-name column; we drop it, remove the `id`, and turn `diagnosis` into a factor with `B` as the reference level.

```{r}
bdiag <- read.csv("bdiag.csv", check.names = FALSE)
bdiag <- bdiag[, nzchar(names(bdiag))]
bdiag$id <- NULL
names(bdiag) <- make.names(names(bdiag))
bdiag$diagnosis <- factor(bdiag$diagnosis, levels = c("B", "M"))
```

## Train/test split

```{r}
n <- nrow(bdiag)
train_idx <- sample.int(n, size = floor(0.75 * n))
train <- bdiag[train_idx, ]
test  <- bdiag[-train_idx, ]
```

## Evaluation helper

```{r}
evaluate <- function(label, truth, prob) {
  pred <- factor(ifelse(prob >= 0.5, "M", "B"), levels = c("B", "M"))
  cm <- table(truth = truth, pred = pred)
  correct <- sum(pred == truth)
  n <- length(truth)
  acc_ci <- binom.test(correct, n)
  acc <- unname(acc_ci$estimate)
  ci  <- acc_ci$conf.int
  auc <- pROC::roc(truth, prob, quiet = TRUE,
                   levels = c("B", "M"), direction = "<")
  auc_ci <- as.numeric(pROC::ci.auc(auc))  # lower, estimate, upper
  cat("\n==", label, "==\n")
  print(cm)
  cat(sprintf("Accuracy: %.4f  (95%% CI %.4f-%.4f)\n", acc, ci[1], ci[2]))
  cat(sprintf("AUC:      %.4f  (95%% CI %.4f-%.4f)\n",
              auc_ci[2], auc_ci[1], auc_ci[3]))
}
```

## 1. Logistic regression on `radius_mean` + `texture_mean`

```{r fit1}
fit1 <- glm(diagnosis ~ radius_mean + texture_mean,
            data = train, family = binomial)
summary(fit1)
prob1 <- predict(fit1, newdata = test, type = "response")
evaluate("Logistic (radius_mean + texture_mean)", test$diagnosis, prob1)
```

## 2. Logistic regression on all features

```{r fit2, warning=FALSE}
fit2 <- glm(diagnosis ~ ., data = train, family = binomial)
prob2 <- predict(fit2, newdata = test, type = "response")
evaluate("Logistic (all features)", test$diagnosis, prob2)
```

## 3. XGBoost on all features

```{r fit3}
feature_cols <- setdiff(names(bdiag), "diagnosis")
x_train <- as.matrix(train[, feature_cols])
x_test  <- as.matrix(test[, feature_cols])
y_train <- as.integer(train$diagnosis) - 1L
y_test  <- as.integer(test$diagnosis)  - 1L

dtrain <- xgb.DMatrix(data = x_train, label = y_train)
dtest  <- xgb.DMatrix(data = x_test,  label = y_test)

params <- list(
  objective = "binary:logistic",
  eval_metric = "logloss",
  eta = 0.1,
  max_depth = 4
)
fit3 <- xgb.train(
  params = params,
  data = dtrain,
  nrounds = 200,
  evals = list(train = dtrain, test = dtest),
  early_stopping_rounds = 20,
  verbose = 0
)
prob3 <- predict(fit3, dtest)
evaluate("XGBoost (all features)", test$diagnosis, prob3)
```

## McNemar test: full logistic vs. XGBoost

McNemar's test compares two classifiers on the same test set by looking at the discordant pairs: cases where one model is correct and the other is wrong. The null hypothesis is that both classifiers have the same error rate.

```{r}
pred2 <- factor(ifelse(prob2 >= 0.5, "M", "B"), levels = c("B", "M"))
pred3 <- factor(ifelse(prob3 >= 0.5, "M", "B"), levels = c("B", "M"))

correct2 <- pred2 == test$diagnosis
correct3 <- pred3 == test$diagnosis

disagreement <- table(
  logistic = factor(correct2, levels = c(TRUE, FALSE), labels = c("correct", "wrong")),
  xgboost  = factor(correct3, levels = c(TRUE, FALSE), labels = c("correct", "wrong"))
)
disagreement

mcnemar.test(disagreement)
```

