---
title: "Modeling and simulation of biosystems"
subtitle: "Case study: river discharge"
author: "Joris Vankerschaver"
header-includes:
  - \useinnertheme[shadow=true]{rounded}
  - \usecolortheme{rose}
  - \setbeamertemplate{footline}[frame number]
  - \usepackage{color}
  - \usepackage{graphicx}
  - \usepackage{siunitx}
  - \usepackage{numprint}
output: 
  beamer_presentation:
    theme: "default"
    keep_tex: true
    includes:
      in_header: columns.tex
---


  
# Model overview
  
---
  
## Setting

\begin{exampleblock}{Streeter-Phelps model}
Use water pollution as water quality monitoring tool. Describes how dissolved oxygen decreases in a river along a certain distance by degradation of biological oxygen demand.
\end{exampleblock}

- Aerobic bacteria gradually remove organic pollution downstream of pollution source
- Reactions
  - Aerobic removal of biochemical oxygen demand
  - Oxygen transfer between atmosphere and water
- Assumption: plug-flow stream
- Simple dynamical model (nonlinear)

## Setting

Typical values for rivers

- Biochemical oxygen demand (BOD)
  - Not polluted: BOD < 1mg/l
  - Mildly polluted: 2mg/l < BOD < 8mg/l
- Dissolved oxygen (DO)
  - Maximal saturation: DO = 12.9mg/l
  - Typical value in freshwater stream: DO $\approx$ 9mg/l
  - Threat to aquatic life: DO < 5mg/l

## Model

\newcommand{\BOD}{\mathsf{BOD}}
\newcommand{\DO}{\mathsf{DO}}

Constant flow rate: location doesn't matter, dynamic model \alert{in time}

\begin{align*}
  \frac{d \BOD}{dt} & = \BOD_{in} - k_1 \BOD \\
  \frac{d \DO}{dt} & = k_2(\DO_{sat} - \DO) - k_1 \BOD
\end{align*}

where

- $\BOD_{in}$: $\BOD$ flux of waste discharge (mg $\cdot$ l$^{-1}$ $\cdot$ min$^{-1}$)
- $\DO_{sat}$: dissolved oxygen concentration at saturation
- $k_1$: deoxygenation rate (min$^{-1}$)
- $k_2$: reaeration rate, rate at which oxygen can be absorbed from the atmosphere (min$^{-1}$)

## Model inputs

- Initial conditions:
  - $\BOD_{t=0} = 7.33$mg/l
  - $\DO_{t=0} = 8.5$mg/l
- Initial model inputs:
  - $\BOD_{in} = 1$mg$\cdot$l$^{-1}$ $\cdot$min$^{-1}$
  - $\DO_{sat} = 8.5$mg$\cdot$l$^{-1}$
  - $k_1 = 0.3$min$^{-1}$ (\alert{unknown})
  - $k_2 = 0.4$min$^{-1}$ (\alert{unknown})

## Model trajectories

```{r, include=FALSE}
source("scripts/03a-parameter-estimation/SP-simulate.R", local = knitr::knit_global())
```

```{r, echo=FALSE, fig.height=6}
plot.exact.trajectories()
```

# Parameter estimation


## Parameter estimation - simplex method

```{r, include=FALSE}
source("scripts/03a-parameter-estimation/SP-estimate.R", local = knitr::knit_global())
```

```{r, echo=FALSE, message=FALSE, fig.height=5}
plot.trajectory("Nelder-Mead")
```

## Parameter estimation - BFGS

```{r, echo=FALSE, message=FALSE, fig.height=5}
plot.trajectory("L-BFGS-B")
```

## Optimal parameter result

```{r, echo=FALSE, message=FALSE, fig.height=6}
plot.fitted.trajectory()
```

# Sensitivity analysis

## Absolute sensitivity functions

Absolute sensitivity of $\DO$ with respect to $k_1$ and $k_2$.

```{r, echo=FALSE, message=FALSE, fig.height=6}
source("scripts/03a-parameter-estimation/SP-sensitivity-absolute.R", local = knitr::knit_global())
```

## Difference exact approximate

```{r, echo=FALSE, message=FALSE, fig.height=6}
source("scripts/03a-parameter-estimation/SP-sensitivity.R", local = knitr::knit_global())
```

## Relative sensitivity functions


```{r, echo=FALSE, message=FALSE, fig.height=6}
source("scripts/03a-parameter-estimation/SP-sensitivity-relative.R", local = knitr::knit_global())
```

## Conclusions

- Note difference in values with absolute sensitivities
- $\DO$ seems more sensitive to $k_2$ than to $k_1$
- Extrema at slightly different time points: information concerning correlation between both parameters
  - Maximal sensitivity at same time point: parameters strongly correlated (impact of change in parameters similar)
- Studying sensitivity very valuable: sensitivity used in many techniques for model analysis  

## Aside: quality of estimation

- DO measurements at different times: measurement error (obtained manually) is 0.05 mg$\cdot$l$^{-1}$
- Estimate simultaneously $k_1$ and $k_2$ (assume all other parameters and initial conditions constant)
- Gives $k_1 = 0.353$min$^{-1}$, $k_2 = 0.389$

## Quality of estimation: FIM

```{r, echo=FALSE, message=FALSE}
source("scripts/03a-parameter-estimation/SP-fisher.R", local = knitr::knit_global())
```

Fisher information matrix:
\[
  \mathsf{FIM} = \sum_{i=1}^N 
    \left(\frac{\partial y}{\partial \theta}(t_i)\right)^T Q_i 
    \left(\frac{\partial y}{\partial \theta}(t_i)\right)
\]

- Measurement noise $\sigma_{DO} = 0.05 \si{mg.l^{-1}}$
- Weight "matrix" in the objective function $Q = \sigma_{DO}^{-2}$.

Gives: 
\[
  \mathsf{FIM} = \left[\begin{matrix}
    \phantom{-} 3.91 \cdot 10^{4} & -2.96 \cdot 10^{4} \\
    -2.96 \cdot 10^{4} & \phantom{-} 4.56 \cdot 10^{5}
  \end{matrix}\right]
\]

## Quality of estimation: confidence intervals

Error covariance matrix:
\[
  \mathsf{C} = \mathsf{FIM}^{-1} = \left[
    \begin{matrix}
      2.69 \cdot 10^{-5} & 1.75 \cdot 10^{-6} \\
      1.75 \cdot 10^{-6} & 2.31 \cdot 10^{-6}
    \end{matrix}
  \right]
\]

95% confidence intervals and covariance:

\begin{align*}
  k_1 & :  0.353 \pm 0.011\\
  k_2 & :  0.389 \pm 0.003\\
  \mathsf{cor}(k_1, k_2) & = 0.22
\end{align*}

## Quantitative analysis

- Calculate $\delta_{rmsq}$ for $\DO$ and different parameters
- Illustration: different measuring schemes for $\DO$:
  - Scheme 1: $t_k = 0 : 0.1 : 2$
  - Scheme 2: $t_k = 0 : 2 : 20$
  - Scheme 3: $t_k = 0 : 2 : 10$
  - Scheme 4: $t_k = 10 : 2 : 20$


```{r, echo=FALSE, message=FALSE}
source("scripts/03a-parameter-estimation/SP-delta-rmsq.R", local = knitr::knit_global())

f <- function(v) {
  round(v, 2)
}
```

\begin{center}
\scriptsize
\begin{tabular}{c|c|c|c|c}
 & \textsf{Scheme 1} & \textsf{Scheme 2} & \textsf{Scheme 3} & \textsf{Scheme 4} \\
\hline
$k_1$ & `r round(d1_k1, 2)` & `r f(d2_k1)` & `r f(d3_k1)` & `r f(d4_k1)` \\
$k_2$ & `r round(d1_k2, 2)` & `r f(d2_k2)` & `r f(d3_k2)` & `r f(d4_k2)` \\
\scriptsize $\DO_{sat}$ & `r round(d1_sat, 2)` & `r f(d2_sat)` & `r f(d3_sat)` & `r f(d4_sat)` \\
\hline
 & $k_2 < k_1 < \DO_{sat}$ & $k_1 < k_2 < \DO_{sat}$ & $k_1 < k_2 < \DO_{sat}$ & $k_1 < k_2 < \DO_{sat}$ \\
\end{tabular}
\normalsize
\end{center}

## GSA: Morris screening: parameter ranges

- Global sensitivity for 6 parameters and initial conditions in designated ranges:

\begin{gather*}
k_1: [0.1; 0.6] \quad k_2: [0.1; 0.6] \quad \DO_{sat} : [10; 12] \\
\BOD_{in} : [0.1; 2] \quad \DO_{t=0}: [6; 10] \quad \BOD_{t=0}: [6; 10]
\end{gather*}


## GSA: Morris screening

```{r, echo=FALSE, message=FALSE, fig.height=4}
source("scripts/03a-parameter-estimation/SP-morris.R", local = knitr::knit_global())
```

Among the selected parameters:

- At time 0, $\DO$ is only sensitive to $\DO_0$
- At time 25, $\DO$ is sensitive to $k_2$ and $\DO_{sat}$.