---
title: "Determinants of pupal color"
format: html
---

Based on Mayekar and Kodandaramaiah (2017), as discussed in Quinn and Keogh, p. 261.

- Data is really unbalanced and not really well separated, hard to get a good classifier.


# EDA

```{r}
library(tidyverse)
```


Explanation of the variables:

- `PC`: Pupal colour
- `H`: Relative humidity
- `SU`: Pupation substrate
- `T`: Time to pupation
- `S`: Sex
- `W`: Pupal weight	

```{r}
pupal <- read.csv("pone.0171482.s005.csv", header = TRUE)
head(pupal)
```

```{r}
glimpse(pupal)
```

## Color against humidity

```{r}
xtabs(~ PC + H, data = pupal)
```

```{r}
pupal_low_humidity <- pupal |>
  filter(H == "low")
```

## Effect of time to gestation on pupal color

```{r}
ggplot(pupal_low_humidity, aes(x = T, y = PC, color = PC)) +
  geom_jitter()
```

```{r}
ggplot(pupal_low_humidity, aes(x = T, y = PC, fill = PC)) +
  geom_boxplot()
```

```{r}
wilcox.test(T ~ PC, data = pupal_low_humidity)
```

```{r}
ggplot(pupal_low_humidity, aes(x = T, y = W, color = PC)) +
  geom_jitter()
```



# Multi-predictor model

```{r}
m <- glm(factor(PC) ~ T + W + S, data = pupal_low_humidity, family = "binomial")
summary(m)
```

# Model building

This is "analysis 2" in the paper.

```{r}
m_interaction <- glm(factor(PC) ~ T + W + S + T:W + T:S + W:S + T:W:S, data = pupal_low_humidity, family = "binomial")
summary(m_interaction)
```

```{r}
library(faraway)
vif(m_interaction)
```
```{r}
cpupal <- pupal_low_humidity |>
  mutate(cT = T - mean(T), cW = W - mean(W)) |>
  select(PC, cT, cW, S)
```

```{r}
m_centered <- glm(factor(PC) ~ cT * cW * S, data = cpupal, family = "binomial")
summary(m_centered)
```

```{r}
m_final <- step(m_centered)
```


```{r}
summary(m_final)
```

Comparison of AIC:
```{r}
AIC(m_final)
AIC(m)
```

Note that we could further simplify the model by removing `W` and `S`

```{r}
m_simple <-  glm(factor(PC) ~ T, data = pupal_low_humidity, family = "binomial")
summary(m_simple)
```

```{r}
AIC(m_simple)
```


```{r}
pupal_low_humidity |>
  mutate(pred = predict(m_simple, type = "response")) |>
  ggplot() +
  geom_line(aes(x = T, y = pred))
```



```{r}
library(caret)

pred_test <- predict(m_simple, pupal_low_humidity, type="response")
class_test <- ifelse(pred_test >= 0.8, "brown", "green")

conf_matrix <- confusionMatrix(as.factor(class_test), as.factor(pupal_low_humidity$PC), positive = "green")
print(conf_matrix)


```

```{r}
library(pROC)

pred_test <- predict(m_simple, pupal_low_humidity, type="response")
roc_logis <- roc(pupal_low_humidity$PC, pred_test)
auc <- round(auc(pupal_low_humidity$PC, pred_test), 4)

ggroc(roc_logis) +
    ggtitle(paste0('ROC Curve ', '(AUC = ', auc, ')'))

```

